### 2018-02-28 thesis/contents/tn_sagit.tex
- sync tn-gp-sagit

The tractography and diffusivity parameters of the trigeminal pathways more central to the brainstem have not been studied, primarily due to methodological limitations. However there are important elements associated with diffusivity abnormalities in TN, including the observation that brainstem diffusivity abnormalities prognosticate a non-responder status and bilateral alterations in trigeminal nerve despite a uniformly unilateral clinical expression of pain. It is therefore relevant to understand the potential TN related abnormalities in more central components of the trigeminal fibers.

\subsubsection{Along-the-tract analysis}
White matter tractography is commonly used for anatomical visualization \cite{Chen2011b}, white matter segmentation \cite{Behrens2003a,Johansen-Berg2005}, and structural connectivity analysis \cite{Cao2013,Wiech2014}. This analysis invovles obtaining diffusivity metrics from tractography after converting streamline bundles into a volumetric spatial mask and measurement of the mean metric from all or parts of the masked voxels \cite{Concha2005,Fitzsimmons2009}. However this approach discards the orientation information in the streamlines. The simplest form of spatial masking is to identify voxels where at least one streamline has passed through it, which often results in identifying voxels with minimal streamline pass-throughs. This is potentially problematic when considering that an average 3-Tesla HARDI DWI acquisition has a voxel dimension of 2 to 3 mm \cite{Neher2015,Wilkins2015}. In white matter structures such as CN V as well as other small pathways, the structure diameter may be fairly small (less than 3mm) and therefore the naive spatial masking approach would result in severe partial volumes that will confound the result.

@@

The global tractography approach is limited to major white matter bundles, and unintuitive for anatomy-specific tasks in a clinical environment, where the anatomy of interest is predetermined. A more focused ROI based delineation approach is therefore preferred in these circumstances. We have developed the Selective Automated Group Integrated Tractography (SAGIT) framework \cite{Chen2016} to automate group-wise tractography delineation, which ensures consistent region-of-interest projection, tractography seeding sampling scheme, and tractography parameter tuning. This technique permits that tracts to be merged into a template space using non-linear deformations. A centroid can be directly defined on the merged tractography bundle, and the diffusivity metrics from multiple subjects can be pooled onto the centroid directly. The extract diffusivity data can then be analyzed by machine learning.

@@

The streamline measurements resembles time-series data, in that it exhibits positional dependency between measurements. Unlike a time-series however, the measurements represent physical locations along a pathway, thus exhibiting bi-directional dependency. Posing the study as a supervised machine learning problem, the feature vector would be a N dimensional vector representing N measures at each division. The GP classifier models the feature vector as a latent GP model, before squashing it to produce a logistic output. Therefore the GP model is able to take the positional dependency of the measurements into account, and is well suited for streamline-based classification.

Gaussian Process is defined \cite{rasmussen2006gaussian} as a collection of random variables, any finite number of which have a joint Gaussian distribution. 
It is a distribution that is defined by a mean function $m(x)$ and covariance function $k(x,x') $ where
\begin{equation}
	\begin{split}
		m(x) &= \mathbb{E}[f(x)], \\
		k(x,x') &= \mathbb{E}[(f(x)-m(x)(f(x')-m(x'))],
	\end{split}
\end{equation}
And the Gaussain processed is defined as
\begin{equation}
f(x) \sim GP(m(x), k(x, x')) 
\end{equation}

The covariance function is termed called kernel function. The radial-basis function kernel (RBF kernel) is widely used in machine learning tasks. GP has previously been explored in neuroimaging for tractography clustering \cite{Wassermann2010} and it is particularly well suited for continuous data streams, such as tractography streamlines.

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{figure-method.pdf}
\caption{Diagram of the processing steps}
\label{fig:GPMethods}
\end{figure}

\subsubsection{Acquisitions}
We studied 36 TN subjects with classic TN (28 females, 8 males; age 60.1$\pm$16 years; right sided pain), and 36 healthy controls (28 females, 8 males, age 42.7$\pm$12 years).  All TN subjects underwent non-invasive Gamma Knife radiosurgery, with imaging prior and six months post treatment. MRI acquisitions including T1 fast-spoiled gradient echo (FSPGR) and diffusion weighted image (DWI) acquisitions. T1 scans were acquired with 1mm slice thickness, in-plane resolution of 0.9375x0.9375mm, slice spacing=1 mm, TE=5.052 ms, TR=11.956 ms, flip angle=20 deg, FOV=240 mm, and matrix=256x256. DWI scans were acquired with 1 B0 scan, 60 gradient directions, 3mm slice thickness and an in-plane resolution of 0.9375x0.9375 mm, b0=1000 s/mm2, TE=88.6 ms, TR=17000 ms, flip angle=90 deg, field of view (FOV)=120 mm, and matrix=128x128.

\subsubsection{Preprocessing}
DWI images were motion/eddy-current corrected, with appropriate correction applied to the b-matrix \cite{Leemans2009}. Region of interests (ROI) for tractography seeding were manually created on a normalized template brain from 42 healthy brains \cite{Chen2016}. The normalized brain template was created using Anatomical Normalization Tools \cite{Avants2010,Avants2011}. We deployed selective automatic group integrated tractography (SAGIT) framework \cite{Chen2016} to automate preprocessing, registration, and tract generation. The SAGIT processing steps are as follows: T1 anatomical images were processed using FreeSurfer \cite{Fischl2004} for cortical and subcortical segmentations. The T1 images for all the subjects were registered to the template using symmetric diffeomorphic registration (SDR) \cite{Avants2008b}. Each individual DWI was registered to their T1 image with SDR, using mean DWI image for DWI-T1 co-registration. Freesurfer segmentation maps were first affine transformed from the Freesurfer space to the T1 space using FMRIB's Linear Image Registration Tool (FLIRT) \cite{Jenkinson2001,Jenkinson2002}, and then transformed using SDR transforms to the individual DWI space for selective ROI generation with SAGIT.

@@

We used constrained spherical deconvolutional tractography (CST) \cite{Tournier2012b} to delineate the three defined anatomical regions. Deterministic and probabilistic streamline algorithms were used to delineate different regions, based on findings in the SAGIT study \cite{Chen2016}. Deterministic CST was applied to delineate CN V and S1 anatomy, while probabilistic (iFOD2) CST \cite{Jeurissen2011b,Tournier2010} was applied to delineate TPT anatomy. The parameters for each were as follows: Deterministic CST was performed with step size = 0.3mm, angle threshold = 45$\deg$, streamline count=200, minimal length = 5 mm, initial cut-off = 0.1, cut-off = 0.1, and with 4th-order Runge-Kutta integration. Probabilistic tractography for S1 projection was performed with step size = 5mm, angle threshold = 45$\deg$, streamline count = 300, max number of stream generation = 1 million, minimal length = 10 mm, maximum lengh = 80 mm, initial cut-off = 0.15, cut-off = 0.15. Probabilistic tractography for TPT project was performed with step size = 1mm, and streamline count = 2000, and maximal length = 60 mm. Tractography was first performed in the native DWI space for each subject.  FA, AD, RD, MD, and GFA were embedded into the native tractography models by sampling the corresponding images maps using the points of the tractography geometry with tri-linear interpolation.

@@

Diffusivity metrics of affected (right side) and unaffected CN V, TPT, and S1 pathways were independently compared with those of controls, where control metrics were the average of both left and right sides. The affected and unaffected sides in TN subjects were also compared with each other. 
Age was regressed out by using the residuals of the linear regression model from the controls group. The diffusion values were subsequently normalized to zero-mean with ranges $[-1, 1]$ for model training. 
Gaussian Process (GP) classification was performed to identify windows that resulted in the highest classification accuracies. Multiple moving windows of length between 5 to 30 were used to scroll through the centroid positions, and a Gaussian Process model was trained for each window (Figure \ref{fig:GPfigure-gp-train}). The machine learning process was performed using Scikit-learn machine learning software framework.
For each window, the subjects were split into train and validation sets using stratified 10-fold cross-validation, and a separate GP classifier was trained for each fold. GP models were initiated with Matern kernel (length scale = 1). The validation set was used to measure prediction accuracy, and the mean accuracy of the 10-fold cross-validation was used to assess the windowed. Classification accuracy greater than 70\% were accepted, and the window with the greatest accuracy was then determined.

@@

\begin{table}[p]
\centering
\small
\csvautotabular{tn_sagit_table.csv}
\caption{List of the best GP models for different diffusivity metrics and centroid windows. Only models with at least 70\% accuracy were accepted. Table is sorted by anatomical region, classifier type and accuracy. Start and end denote the beginning and end of the window subdivision positions along the centroid. Length is the number of subdivisions covered by the window. Precision and f1 scores are provided for reference}
\label{table:GP}
\end{table}

\section{Discussion}
We have demonstrated the use of a fully-automated, end-to-end tractography and machine learning platform that is capable of discriminate TN from white-matter diffusivity with greater than 80\% accuracy strictly from white-matter measures. Our work is in agreement with previous findings of TN diffusivity changes along the CN V, yet the present technique offers much higher regional specificity. We also examined TPT and S1 projections of the trigeminal sensory pathway. The study is the first instance of machine learning discrimination of TN vs. controls with specific white matter analysis, using group-wise merged along-the-track tractography.

@@

The TPT left-verus-right FA/RD/GFA classifiers demonstrated very high accuracy (>90\%), with overlapping windows in the ascending segment. The TPT merged tractography delineations showed that decussation pathway tractography is possibly asymmetric. The asymmetry of this decussation has not been demonstrated before. It is not clear whether this represents the anatomical arrangement of these fibers, or whether decussations are always purely symmetrical (akin to a perfectly shuffled deck of cards). Further assessment of other decussation points will be helpful to understand this process.

@@

The DWI images were acquired at 0.9375$\times$0.9375$\times$3 $mm\textsuperscript{3}$ voxel resolution. The acquisition was a compromise to allow better in-plane resolution to image the trigeminal nerve, due to the limits of the 3T GE HDx MRI with 8-channels head coil. Our analysis is internally consistent, and SAGIT was previously validated on a identical sequence \cite{Chen2016,Chen2016a}, therefore we believe the approach is valid.

>>>>

The tractography and diffusivity parameters of the trigeminal pathways more central to the brainstem have not been studied, primarily due to methodological limitations. However there are important elements associated with diffusivity abnormalities in TN, including the observation that brainstem diffusivity abnormalities prognosticate a non-responder status \cite{Hung2017} and bilateral alterations in trigeminal nerve despite a uniformly unilateral clinical expression of pain \cite{Miller2009}. It is therefore relevant to understand the potential TN related abnormalities in more central components of the trigeminal fibers.

\subsubsection{Along-the-tract analysis}
White matter tractography is commonly used for anatomical visualization \cite{Chen2011b}, white matter segmentation \cite{Behrens2003a,Johansen-Berg2005}, and structural connectivity analysis \cite{Cao2013,Wiech2014}. This analysis invovles obtaining diffusivity metrics from tractography after converting streamline bundles into a volumetric spatial mask and measurement of the mean metric from all or parts of the masked voxels \cite{Concha2005,Fitzsimmons2009}. However this approach discards the orientation information in the streamlines. The simplest form of spatial masking is to identify voxels where at least one streamline has passed through it, which often results in identifying voxels with minimal streamline pass-throughs. This is potentially problematic when considering that an average 3-Tesla HARDI DWI acquisition has a voxel dimension of 2 to 3 mm \cite{Neher2015,Wilkins2015}. In white matter structures such as CN V as well as other small pathways, the structure diameter may be fairly small (less than 3 mm) and therefore the naive spatial masking approach would result in severe partial volumes that will confound the result.

@@

The global tractography approach is limited to major white matter bundles, and unintuitive for anatomy-specific tasks in a clinical environment, where the anatomy of interest is predetermined. A more focused ROI based delineation approach is therefore preferred in these circumstances. We have developed the Selective Automated Group Integrated Tractography (SAGIT) framework \cite{Chen2016} to automate group-wise tractography delineation, which ensures consistent region-of-interest (ROI) projection, tractography seeding sampling scheme, and tractography parameter tuning. This technique permits tracts to be merged into a template space using non-linear deformations. A centroid can be directly defined on the merged tractography bundle, and the diffusivity metrics from multiple subjects can be pooled onto the centroid directly. The extact diffusivity data can then be analyzed by machine learning.

@@

The streamline measurements resembles time-series data, in that it exhibits positional dependency between measurements. Unlike a time-series however, the measurements represent physical locations along a pathway, thus exhibiting bi-directional dependency. Posing the study as a supervised machine learning problem, the feature vector would be an N-dimensional vector representing N measures at each division. The GP classifier models the feature vector as a latent GP model, before squashing it to produce a logistic output. Therefore the GP model is able to take the positional dependency of the measurements into account, and is well suited for streamline-based classification.

Gaussian Process is defined \cite{rasmussen2006gaussian} as a collection of random variables, any finite number of which have a joint Gaussian distribution. 
It is a distribution that is defined by a mean function $m(x)$ and covariance function $k(x,x') $ where
\begin{equation}
	\begin{split}
		m(x) &= \mathbb{E}[f(x)], \\
		k(x,x') &= \mathbb{E}[(f(x)-m(x))(f(x')-m(x'))],
	\end{split}
\end{equation}
and the Gaussain processed is defined as
\begin{equation}
f(x) \sim GP(m(x), k(x, x')) 
\end{equation}

The covariance function is also called the kernel function. The radial-basis function kernel (RBF kernel) is widely used in machine learning tasks. GP has previously been explored in neuroimaging for tractography clustering \cite{Wassermann2010} and it is particularly well suited for continuous data streams, such as tractography streamlines.

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\linewidth]{figure-method.pdf}
\caption{Diagram of the overall processing steps. The details of SAGIT processing steps can be found in \protect\cite{Chen2016}. Details of the centroid identification can be found in Figure \ref{fig:GPfigure-method-centroid}. The details for machine learning process can be found in Figure \ref{fig:GPfigure-gp-train}. }
\label{fig:GPMethods}
\end{figure}

\subsubsection{Acquisitions}
We studied 36 TN subjects with classic TN (28 females, 8 males; age 60.1$\pm$16.0 years; right sided pain), and 36 healthy controls (28 females, 8 males, age 42.7$\pm$12.0 years).  All TN subjects underwent non-invasive Gamma Knife radiosurgery, with imaging prior and six months post treatment. MRI acquisitions including T1 fast-spoiled gradient echo (FSPGR) and diffusion weighted image (DWI) acquisitions. 
T1 scans were acquired with 1mm slice thickness, in-plane resolution of 0.9375x0.9375 mm, with slice spacing = 1 mm, TE = 5.052 ms, TR = 11.956 ms, flip angle = $20^\circ$, FOV = 240 mm, and matrix = 256x256. 
DWI scans were acquired with 1 B0 scan, 60 gradient directions, 3 mm slice thickness and an in-plane resolution of 0.9375x0.9375 mm, where b0 = 1000 s/$mm^2$, TE = 88.6 ms, TR = 17000 ms, flip angle = $90^\circ$, field of view (FOV) = 120 mm, and matrix = 128x128.

\subsubsection{Preprocessing}
DWI images were motion/eddy-current corrected, with appropriate correction applied to the b-matrix \cite{Leemans2009}. ROI for tractography seeding were manually created on a normalized template brain from 42 healthy brains \cite{Chen2016}. The normalized brain template was created using Anatomical Normalization Tools \cite{Avants2010,Avants2011}. We deployed selective automatic group integrated tractography (SAGIT) framework \cite{Chen2016} to automate preprocessing, registration, and tract generation. The SAGIT processing steps are as follows: T1 anatomical images were processed using FreeSurfer \cite{Fischl2004} for cortical and subcortical segmentations. The T1 images for all the subjects were registered to the template using symmetric diffeomorphic registration (SDR) \cite{Avants2008b}. Each individual DWI was registered to their T1 image with SDR, using mean DWI image for DWI-T1 co-registration. Freesurfer segmentation maps were first affine transformed from the Freesurfer space to the T1 space using FMRIB's Linear Image Registration Tool (FLIRT) \cite{Jenkinson2001,Jenkinson2002}, and then transformed using SDR transforms to the individual DWI space for selective ROI generation with SAGIT.

@@

We used constrained spherical deconvolutional tractography (CST) \cite{Tournier2012b} to delineate the three defined anatomical regions. Deterministic and probabilistic streamline algorithms were used to delineate different regions, based on findings in the SAGIT study \cite{Chen2016}. Deterministic CST was applied to delineate CN V and S1 anatomy, while probabilistic (iFOD2) CST \cite{Jeurissen2011b,Tournier2010} was applied to delineate TPT anatomy. The parameters for each were as follows: Deterministic CST was performed with step size = 0.3 mm, angle threshold = $45^\circ$, streamline count = 200, minimal length = 5 mm, initial cut-off = 0.1, cut-off = 0.1, and with 4th-order Runge-Kutta integration. Probabilistic tractography for S1 projection was performed with step size = 5mm, angle threshold = $45^\circ$, streamline count = 300, max number of stream generation = 1 million, minimal length = 10 mm, maximum length = 80 mm, initial cut-off = 0.15, cut-off = 0.15. Probabilistic tractography for TPT project was performed with step size = 1 mm, and streamline count = 2000, and maximal length = 60 mm. Tractography was first performed in the native DWI space for each subject.  FA, AD, RD, MD, and GFA were embedded into the native tractography models by sampling the corresponding images maps using the points of the tractography geometry with tri-linear interpolation.

@@

Diffusivity metrics of affected (right side) and unaffected CN V, TPT, and S1 pathways were independently compared with those of controls, where control metrics were the average of both left and right sides. The affected and unaffected sides in TN subjects were also compared with each other. 
Age was regressed out by using the residuals of the linear regression model from the controls group. The diffusion values were subsequently normalized to zero-mean with ranges $[-1, 1]$ for model training. 
Gaussian Process (GP) classification was performed to identify windows that resulted in the highest classification accuracies. Multiple moving windows of length between 5 to 30 were used to scroll through the centroid positions, and a Gaussian Process model was trained for each window (Figure \ref{fig:GPfigure-gp-train}). The machine learning process was performed using Scikit-learn machine learning software framework.
For each window, the subjects were split into train and validation sets using stratified 10-fold cross-validation, and a separate GP classifier was trained for each fold. GP models were initiated with Matern kernel (length scale = 1). The validation set was used to measure prediction accuracy, and the mean accuracy of the 10-fold cross-validation was used to assess the windows. Classification accuracy greater than 70\% were accepted, and the window with the greatest accuracy was then determined.

@@

\begin{table}[p]
\centering
\small
\csvautotabular{tn_sagit_table.csv}
\caption{List of the best GP models for different diffusivity metrics and centroid windows. Only models with at least 70\% accuracy were accepted. Table is sorted by anatomical region, classifier type (R denotes right (affected) side versus controls; L denotes left (unaffected) side versus controls; LVR denotes left-versus-right ) and accuracy. Start and end denote the beginning and end of the window subdivision positions along the centroid. Length is the number of subdivisions covered by the window. Precision and f1 scores are provided for reference}
\label{table:GP}
\end{table}

\section{Discussion}
We have demonstrated the use of a fully-automated, end-to-end tractography and machine learning platform that is capable of discriminate TN from white-matter diffusivity with greater than 80\% accuracy strictly from white-matter measures. Our work is in agreement with previous findings of TN diffusivity changes along the CN V, yet the present technique offers much higher regional specificity, as well as examining the TPT and S1 projections of the trigeminal sensory pathway. The study is the first instance of machine learning discrimination of TN vs. controls with specific white matter analysis, using group-wise merged along-the-track tractography.

@@

The TPT left-verus-right FA/RD/GFA classifiers demonstrated very high accuracy (\textgreater 90\%), with overlapping windows in the ascending segment. The TPT merged tractography delineations showed that decussation pathway tractography is possibly asymmetric. The asymmetry of this decussation has not been demonstrated before. It is not clear whether this represents the anatomical arrangement of these fibers, or whether decussations are always purely symmetrical (akin to a perfectly shuffled deck of cards). Further assessment of other decussation points will be helpful to understand this process.

@@

The DWI images were acquired at 0.9375$\times$0.9375$\times$3 $mm\textsuperscript{3}$ voxel resolution. The acquisition was a compromise to allow better in-plane resolution to image the trigeminal nerve, due to the limits of the 3T GE HDx MRI with 8-channels head coil. Our analysis is internally consistent, and SAGIT was previously validated on a identical sequence \cite{Chen2016,Chen2016a}, highlighting the validity of this approach.
